---
title: "testing2_summarizing_for_time_series"
format: html
---

## Set up

```{r}
library(tidyverse)
library(janitor)
library(readxl) # reading excel files
library(ggeffects) # getting model predictions
library(MuMIn) # model selection
library(vegan)  #for diversity index
library(hillR) #for calculating hill numbers


# read in your data here
orig_data <- read_xlsx("LENC_Database_Ind.Assign_2023-2024_Good.xlsx")


```

## Cleaning

```{r}

clean_name_data <- orig_data |> 
  clean_names() |> 
  mutate(location_group = if_else(percent_build_up <= 50, "Rural", "Urban"))


rural_bird_data <- clean_name_data |> 
  filter(location_group == "Rural") |> #select only the rows with "rural" as the location_group
  select(19:96) |> #select only the bird species columns
  mutate(across(where(is.numeric), ~ replace_na(., 0))) |> #replace any NA with 0's 
  select(where(~ sum(.) > 0))

rural_bird_matrix <- as.matrix(rural_bird_data)

urban_bird_data <- clean_name_data |> 
  filter(location_group == "Urban") |> 
  select(19:96) |> 
  mutate(across(where(is.numeric), ~ replace_na(., 0))) |> 
  select(where(~ sum(.) > 0))



```


## Summarizing and Finding Hill Numbers

```{r}
#goal is to summarize all the data by different groups, this group will be bird surveys grouped by built up percent cut off

hill_rual_bird <- hill_taxa(comm = rural_bird_data, q = 1)
hill_rural_bird_df <- as.data.frame(hill_rual_bird)

hill_urban_bird <-  hill_taxa(comm = urban_bird_data, q = 1)
hill_urban_bird_df <- as.data.frame(hill_urban_bird)


combined_summary <- cross_join(hill_rural_bird_df, hill_urban_bird_df) #join each survey type summaries by location



```

❓ How Would You Know Which q to Use?

Use q = 0 if you're interested in species richness only (ignores abundance).

Use q = 1 for a balanced measure sensitive to abundance but not dominated by it.

Use q = 2 to emphasize dominant species and reduce weight of rare ones.

Plotting diversity profiles over a range of q-values (e.g. 0–5) can help you understand diversity patterns more fully.





grandson: (playing with pink shift, didn't list ones where pink shift wasnt there)

berlin 30 jan friday
utrecht 1 feb sunday
muster (germany) 6 feb friday
copenhagen 7 feb saturday
wiesbaden (germany) 21 feb saturday
birmingham (uk) 28 feb saturday
sofia (bulgaria) 7 march saturday


callous daoboys: (no knives & Love Rareley <- only in UK)

hamburg (germany) 6 feb friday
berlin (germany) 7 feb saturday
lyon (france) 14 feb saturday
brighton (uk) 20 feb friday
southhampton (uk) 21 feb saturday
bristol (uk) 27 feb friday
london (uk) 28 feb saturday








